A Global Picture of Hunger

### 2020-11-15 docs/modeling_fies.tex
new model outputs

\title{A Global Picture of Food Insecurity and Hunger}

\author[1,2,*]{Matthew Cooper}
\author[2,3]{Benjamin Müller}
\author[4]{Carlo Cafiero}
\author[2,5]{Juan Carlos Laso Bayas}
\author[2,6]{Homi Kharas}

\affil[1]{T.H. Chan School of Public Health, Harvard University}
\affil[2]{The World Data Lab, Vienna, Austria}
\affil[3]{Department of Economics, Vienna University of Economics and Business}
\affil[4]{The Food and Agricultural Organization of the United Nations}
\affil[5]{The International Institute for Applied Systems Analysis}
\affil[6]{The Brookings Institute}
\affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}

\maketitle
\begin{abstract}
We modeled levels of of food security at a subnational level from 2010 to 2030 using microdata from 75 countries collected by the FAO and Gallup World Poll for the Voices of the Hungry project.  We find significant heterogeneity in levels of food security around the world, ranging from parts of the global north with less than 1\% of the population food insecure to parts of the global south where as much as 3 out of 4 people are severely food insecure.  Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that food insecurity has increased slightly over the past several years, but under middle-of-the-road assumptions for development and population change, food insecurity will decline throughout the 2020s.  This global decrease is largely driven by trends in South and East Asia, and some parts of the world, particularly sub-Saharan Africa, are projected to show increases in the prevalence of food insecurity throughout the next decade. Our model's forecast of decreasing food insecurity through the 2020s contrasts with predictions from the FAO that other metrics of hunger will increase.  This is the first global picture at a subnational level of the Food Insecurity Experience Scale, the food security metric most indicative of the lived experience of hunger.
\end{abstract}

\section{Main}
Food security is a critical component of human flourishing, which is why the second Sustainable Development Goal (SDG 2) is “Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of moderate to severe food insecurity in the population based on the Food Insecurity Experience Scale, or FIES. The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience hunger and conditions that produce hunger. Other indicators of food insecurity, such as FAO's most commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories consumed per capita in a population. These do not capture the actual lived exposure of individuals to food shortages.

Because the FIES is based on responses to polls, the data is limited to 75 countries where the polls have been conducted and microdata has been vetted and released by the FAO. The current practice is to assign the regional average prevalence rate for countries without survey data in order to compile regional and global aggregates. This is not very satisfactory. In addition, because the FIES is a relatively new metric, it does not have a sufficiently long time series to estimate trends with any degree of precision. To understand whether SDG 2 is being met, one needs to estimate food insecurity outcomes in 2030.

The contribution of this paper is to fill these two gaps of coverage and forecasts by developing a model of FIES outcomes that is based on covariates that are available for places where actual polling data is not available, and that can be forecast to 2030 to get a sense of what food insecurity at that time might look like. We also take advantage of the within-country spatial distribution of FIES reports to push the analysis beyond country level to a sub-national level, that is administratively once removed from the national level (Admin-1). To our knowledge, this is the first effort to model the FIES outcomes at that scale.

Food security has traditionally been difficult to measure, and this has led to incomplete or inaccurate pictures of global hunger.  Metrics of macro-health, such as anthropometry and mortality rates are correlated broadly with food insecurity and have been used for many years to monitor human well-being \citep{Puffer1973, Habicht1974}.  However, these metrics are confounded with other determinants of health such as infectious disease and are not meaningful at the scale of individuals or households \citep{Perumal2018}.  Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they only make rough estimates of how accessible food is to the general population, and food insecurity can certainly occur in the absence of food availability decline \citep{Sen1983}.  Moreover, these metrics are very sensitive to incorrect estimates of crop yields and food reserves at a national scale.  Thus, global estimates of hunger and food insecurity based on these metrics carry forward similar flaws.

@@

Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the UN \cite{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.  The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.  Using a Rash model \citep{Cafiero2018}, each individual in the survey is given a severity score that maximizes the likelihood of observing he responses given to the 8 questions asked in the survey.  Once individuals are ranked in this fashion, thresholds for moderate and severe food insecurity are set that describe people's experiences, and the prevalences can be computed. A moderately food insecure adult is likely to have compromised on food quality and variety, as a result of lack of income, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally. A severely food insecure adult is likely to have run out of food and has gone an entire day without eating at times during the year.  Because the model estimates the population over a threshold on a univariate food insecurity scale, the percentage of people over the threshold for moderate food insecurity includes people that are also over the threshold for severe food insecurity.

Since 2014, in partnership with Gallup World Poll, the FAO administered the FIES in surveys around the world. The FAO uses this data to estimate the prevalence of food insecurity, and reports national-level estimates of the percentage of the population over the thresholds for moderate and severe food insecurity.  Drawing on individual-level data from these microdatasets in 75 countries, we used machine learning to estimate levels of food insecurity around the world at the subnational level.  The model is then used to infer levels of food insecurity around the world at the subnational level and to forecast food insecurity to the year 2030.

\section{Results}
\subsection{Food Insecurity over Time}
Globally, in the year 2020, we estimate that 829 million people are above the threshold for severe food insecurity, and an additional 1.5 billion people are above the threshold for moderate food insecurity (See Fig. \ref{fig:timeseries}). These numbers are higher than the estimates in the 2019 FAO State of Food Insecurity report, which estimated about 700 million severely food insecure people and an additional 1.31 billion people experiencing at least moderate food insecurity in 2018.

While the total number of people experiencing moderate and severe food insecurity held roughly constant for the 2010s and even increased in latter part of that decade, our model predict that it will decline throughout the 2020s. However, this overall decline will not be uniform across the world. Our model expects that, in most of sub-Saharan Africa, moderate food insecurity will increase throughout the 2020s and severe food insecurity will plateau, although some countries, such as Kenya, Ethiopia, and Ghana, will see improvements. Food insecurity will fall in most other world regions, particularly South Asia and East Asia \& the Pacific.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
  \caption{Number of people over the thresholds for moderate and severe food insecurity, by UN SDG regions over time, with a 3-year smoothing.  Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.  Panel (B) shows regional totals compared against each other over time.}
  \label{fig:timeseries}
\end{figure}

Examining the rate of food insecurity in a population, rather than the total number of food insecure people (See Fig. \ref{fig:rates}, our model finds the most progress has been made in East \& Southeast Asia, while Central \& Southern Asia is expected to see improvements in food insecurity throughout the 2020s.  Other poor and middle income areas, including much of Africa, the Middle East and Latin America, have made only slight progress and will continue to make slight progress between 2010 and 2030.  In high-income countries in Europe and North America, rates of people in at least moderate food insecurity move little, and rates of severe food insecurity are expected to increase slightly.  Overall, the world has made steady but slow progress on reducing the rate of food insecurity.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/Rates.pdf}
  \caption{Percentage of the population over the threshold for moderate and severe food insecurity, by UN SDG regions over time, with a 3-year smoothing.}
  \label{fig:rates}
\end{figure}

\subsection{Food Insecurity over Space}
We found substantial heterogeneity in the global distribution of severe food insecurity (See Fig. \ref{fig:map}).  For the year 2020, mainland sub-Saharan Africa is the continent with the highest rates of severe food insecurity, with at least 15\% of people over the threshold for severe food insecurity in at last one subnational area in every country except Gabon and Equatorial Guinea.  Outside of Africa, serious pockets of severe food insecurity also occur in Venezuela, Syria, Papua New Guinea, Yemen, and Afghanistan.  In many large middle-income countries, severe food insecurity is also quite prevalent, with rates between 10-15\% in areas like northern Brazil, many central Asian and middle-eastern countries, and India and Indonesia.

The experience of at least moderate food insecurity is quite common in many parts of the world.  In 2020, much of Africa, south and southeast Asia, and parts of Latin America had over 50\% of the population living with moderate food insecurity.  Even in countries widely considered to be developed, such as Australia, the United States, and parts of eastern and southern Europe, over 10\% of the population is above the threshold for moderate food insecurity.

\begin{landscape}
\begin{figure}[h]
  \centering
  \includegraphics{img/FullMap.png}
  \caption{Modeled spatial distribution of moderate and severe food insecurity for the years 2010, 2020, and 2030.}
  \label{fig:map}
\end{figure}
\end{landscape}

\subsection{Important Predictors of Food Insecurity}
Overall, our model was able to predict food insecurity with high accuracy in the countries that had training data ($R^2 = 0.9981$ and $R^2 = 0.9976$ for moderate and severe food insecurity, respectively).  Examining how permuting a variable affects the predictive accuracy of the model can give an idea of how important that variable is for predicting food insecurity.

For both moderate and severe food insecurity, the most important variable for predicting food security levels was the Poverty Headcount Index, followed by the rate of stunting.  For predicting the rate of people with at least moderate food insecurity, GDP per capita was quite important.  However, for the rate of severe food insecurity, the country's Gini coefficient was more important compared to moderate food insecurity.

@@

\section{Discussion}
This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts out to the year 2030.  We found that, while food insecurity has been increasing and has likely worsened in 2020 with economic downturn associated with the novel coronavirus, our models predict that number of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in south and east Asia.  Nevertheless, many regions, most notably sub-Saharan Africa, are expected to see overall increases in the number of food insecure people.

Our variable importance metrics show that economic conditions are the largest driver of global variation in rates of food insecurity.  For both moderate and severe food insecurity, the poverty headcount index, the Gini coefficient, and GDP per capita were major predictors of the rate of food insecurity.  The rate of stunting, an indicator of both sanitary conditions and chronic, long-term hunger was also an important predictor of food insecurity.  However, other indicators of sanitary conditions, such as the rates of wasting and malaria mortality were less important predictors of food insecurity.  Geographic and climate factors, such as urban percentage, topographic ruggedness and mean annual temperature and precipitation were not important predictors of food insecurity at a global scale.  Finally, while economic conditions are widely associated with food insecurity at a global scale, the fact that Europe and North America is predicted to see little improvement in food insecurity outcomes between 2010 and 2030, in spite of economic growth and continued reductions in poverty, suggests that there is an upper limit to the extent to which increasing economic development can improve food security outcomes.

These results contrast with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.  This is likely due in part to our differing modeling approaches. While the SOFI report extrapolated recent trends in undernourishment and stunting, we modeled food insecurity based on interactions between projections in factors like GDP per capita, population, and the prevalence of stunting. Additionally, our diverging predictions for different SDG 2 indicators are not necessarily mutually exclusive: it is theoretically possible for undernourishment to increase while food insecurity, as measured by the FIES, decreases.  The discrepancies between our projections illustrate both how hunger is a complex phenomenon that is resistant to simple quantification, as well as the challenges in forecasting the future, where different modeling approaches will yield different predictions.

When making predictions in countries where we did not have primary data, we are assuming that the relationship between variables like poverty, stunting, inequality, and hunger are the same in all locations.  We tested the accuracy of these cases by using ten-fold cross validation at the country level, and we found reasonable accuracy ($R^2 = 0.778$, MAE = 0.073 for moderate and $R^2 = 0.696$, MAE = 0.044 for severe food insecurity).  This suggestions that our model explains between 70-78\% of the variation in food security outcomes and is on average, within 4-7\% in its estimates of rates of food insecurity, when applied to countries outside of the training set.

Additionally, when making predictions into the future, we are assuming that long-term patterns of demographic change, urbanization, and development will maintain their long-held trajectories for the next decade.  At a global scale, this is almost certainly the case: it is highly unlikely that rates of fertility or economic growth will shift suddenly and globally.  Nevertheless, at more local scales, sudden crises can lead to severe and unforeseeable increases in food insecurity, as crises in the previous decade in Yemen, Syria and Venezuela have shown.  Thus, while our model's predictions at a global scale are very likely to bear out, there will probably be individual countries and areas where our forecasts end up being off.

While we found that food insecurity will decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, only expects the number of people experiencing moderate food insecurity to fall by 12\% and the number of people experiencing severe food insecurity to fall by 20\%.  This still leaves billions of people eating less than they should and nearly half a billion people going entire days without eating a decade from now, a number that falls far short of the SDG 2 goal of ending hunger.  Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, or obesity.  Thus, while expected trends in correlates of food security give cause for optimism, there is still significant work to be done in pursuit of SDG 2.

\section{Methods}

%do we need to add a section where we state which fies data we use? like countries and years?

\subsection{Disaggregation}
The data on the FIES collected by Gallup records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.  Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.  We used a similar methodology, except we calculated post-stratification weights at a subnational level.  We used year-specific subnational estimates of population percentages by gender, age, urbanization, and, where available, wealth, and created a separate set of weights for each subnational area across all individuals.  For age and gender, we used gridded data from WorldPop \citep{Tatem2017}, for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}, and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}, where available.  This methodology rests on the mild assumption that subnational differences in rates of food insecurity are more due to factors related to demographics, urbanization, and wealth than to other unobservable factors.

\subsection{Covariates}
We model subnational rates of severe and moderate-to-severe food insecurity as a function of several covariates related to human development, food security, health, infrastructure, income levels and distribution, and the environment (See Table \ref{tab:covars}).  We aimed for covariates that had subnational spatial resolution and peer-reviewed projections available through the year 2030.   For projections based on the Shared Socioeconomic Pathways (SSP) framework \citep{oneill2014new}, we used the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on Representative Concentration Pathway (RCP) 6.0 \citep{van2011representative}.

Many of the variables required harmonizing subnational historic data with projections at the national level.  This included mean years of schooling, GDP per capita, as well as for projections of population. We used observed trajectories in the historical distribution of variables among subnational areas within a country to estimate the future distribution of GDP, population, and schooling among subnational areas and disaggregate national-level future projections.

In cases of health variables that have shown long-term trends and represented a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these variables for 2030. The involves taking the rate of change between each pair of years in the dataset, and then taking the mean rate of change over the period for which data is available, giving greater weight to more recent years, and applying that mean rate of change to estimate future levels.  This method has been used to estimate rates of stunting and wasting in 2025 \citep{Local2020}; we simply use the same method to make forecasts to the year 2030.

For the climate variables temperature and precipitation, we combined historical observations with an ensemble of four bias-corrected simulations of the future climate from the Inter-Sectoral Impact Model Intercomparison Project (ISIMIP) \citep{warszawski2014inter}.

Finally, we adjusted several variables to account for the effects of COVID-19, including GDP per capita using June 2020 estimates from the World Bank \citep{prospects2020}, stunting and wasting using August 2020 estimates published in \textit{The Lancet} \citep{headey2020impacts}, as well as the poverty headcount index \cite{Cuaresma2018}.  For a detailed overview of the steps involved in processing and preparing each covariate, see the Appendix.

>>>>

\title{A Global Picture of Hunger}

\author[1,2,*]{Matthew Cooper}
\author[2,3]{Benjamin Müller}
\author[4]{Carlo Cafiero}
\author[2,5]{Juan Carlos Laso Bayas}
\author[5,6]{Jesús Crespo Cuaresma}
\author[2,7]{Homi Kharas}

\affil[1]{T.H. Chan School of Public Health, Harvard University}
\affil[2]{The World Data Lab, Vienna, Austria}
\affil[3]{Department of Economics, Vienna University of Economics and Business}
\affil[4]{The Food and Agricultural Organization of the United Nations}
\affil[5]{The International Institute for Applied Systems Analysis}
\affil[6]{Wittgenstein Centre For Demography and Global Human Capital}
\affil[7]{The Brookings Institute}
\affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}

\maketitle
\begin{abstract}
Using microdata from 75 countries collected by the FAO and Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the \href{https://worldhunger.io}{World Hunger Clock}.  This is the first global picture at a subnational level of the Food Insecurity Experience Scale, the food security metric most indicative of the lived experience of hunger.  We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population food insecure in some high-income countries to the developing world where over half the population is severely food insecure.  Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of food insecurity have been steadily declining, with large observed and forecasted gains in lower and lower-middle income countries, and decreases in total number of severely food insecure people.  However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, will continue to increase through the end of the 2020s.  Overall, we conclude that global gains have been incremental, and current trends in development and demographic change will still leave a large share of the world's population experiencing hunger by 2030.  \end{abstract}

\section{Main}
Food security is a critical component of human flourishing and its importance as a global policy objective is reflected in the second Sustainable Development Goal (SDG 2), ``Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of moderate to severe food insecurity in the population based on the Food Insecurity Experience Scale, or FIES. The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience hunger and conditions that produce hunger. Other indicators of food insecurity, such as FAO's most commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories available per capita in a population. These do not directly capture the actual lived exposure of individuals to food shortages.

The FIES is based on responses to polls and the data are limited to 75 countries where these have been conducted and microdata has been vetted and released by the FAO. The current practice is to assign the average regional prevalence rate for countries without survey data in order to compile regional and global aggregates. In addition, since the FIES is a relatively new metric, it does not have a sufficiently long time series to assess time trends with a reasonable degree of precision. To understand whether SDG 2 is being met, estimates of food insecurity outcomes in 2030 are required.

This paper contributes to shifting the research frontier by filling these two gaps in terms of data coverage and forecasts by developing a statistical model of FIES outcomes based on covariates that are available for territories where actual polling data are not available. Such a model can be employed to obtain forecasts to 2030 and assess likely future trends in food insecurity. We take advantage of individual-level characteristics collected in FIES surveys to push the analysis beyond country level to a sub-national level. To our knowledge, this is the first effort to model the FIES outcomes at that scale.

Food security has traditionally been difficult to measure, and this has led to an incomplete or inaccurate assessment of global hunger. Metrics of macro-health, such as anthropometric measures and mortality rates are correlated broadly with food insecurity and have been used for many years to monitor human well-being \citep{Puffer1973, Habicht1974}.  However, these metrics are affected by other determinants of health such as the occurrence of infectious diseases, and are not meaningful at the scale of individuals or households \citep{Perumal2018}.  Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they only make rough estimates of how accessible food is to the general population, and can elide populations that food insecure due to poor access, even when aggregate food availability is high \citep{Sen1983}.  Moreover, these metrics are very sensitive to incorrect estimates of crop yields and food reserves at a national scale.  Thus, global estimates of hunger and food insecurity based on these metrics carry forward similar flaws.

@@

Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the UN \cite{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.  The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.  Using a Rasch model \citep{Cafiero2018}, each individual in the survey is given a severity score that maximizes the likelihood of observing the responses given to the 8 questions asked in the survey.  Once individuals are ranked in this fashion, thresholds for moderate and severe food insecurity are set that describe people's experiences, and the prevalence of such food insecurity events can be computed. A moderately food insecure adult is likely to have compromised on food quality and variety as a result of lack of income, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally. A severely food insecure adult is likely to have run out of food and has gone an entire day without eating at times during the year.  Because the model estimates the population over a threshold on a univariate food insecurity scale, the percentage of people over the threshold for moderate food insecurity includes people that are also over the threshold for severe food insecurity.

Since 2014, in partnership with Gallup World Poll, the FAO administered the FIES in surveys around the world. The FAO uses this data to estimate the prevalence of food insecurity, and reports national-level estimates of the percentage of the population over the thresholds for moderate and severe food insecurity.  Drawing on individual-level data from these microdatasets in 75 countries, we use machine learning methods to estimate levels of food insecurity globally at the subnational level.  The model is then used to forecast food insecurity to the year 2030 under a middle-of-the-road scenario.

\section{Results}
\subsection{Food Insecurity over Time}
Globally, in the year 2020, we estimate that 820 million people are above the threshold for severe food insecurity, and an additional 2.5 billion people are above the threshold for moderate food insecurity (See Fig. \ref{fig:timeseries}). These numbers are higher than the estimates in the 2019 FAO State of Food Insecurity report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \ref{sofi2020}.

Under realistic scenario assumptions, we find divergent trends in the number of moderate and severely food insecure people around the world, with the number of people at least moderately food insecure increasing and the number of people severely food insecure declining.  However, for both indicators, these changes are very slight: from 2010 to 2030, the number of moderately food insecure people will increase 2\%, and the number of severely food insecure people will decrease 8\%.  The increases in moderate food insecurity have been in lower and lower-middle income countries, while the increases in severe food insecurity are entirely from lower income countries.  For both the moderate and severe food insecurity thresholds, COVID-19 has had a clearly discernible impact, particularly in middle-income countries.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
  \caption{Number of people over the thresholds for moderate and severe food insecurity, by World Bank income groups as of 2020.  Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.  Panel (B) shows regional totals compared against each other over time.}
  \label{fig:timeseries}
\end{figure}

Examining the rate of food insecurity in a population, rather than the total number of food insecure people (See Fig. \ref{fig:rates}, we find clear improvements in all except high income countries.  Countries that are currently upper-middle income, like China and Brazil, saw large gains in improving food insecurity rates throughout the 2010s, and will continue to see somewhat diminished gains in improving food security throughout the 2030s.  Lower and lower-middle income countries, on the other hand, saw more modest gains in the 2010s but are forecasted to see greater gains in reducing the rate of food insecurity in 2020s.  In spite of this progress in the bottom three income quartiles, in high-income countries, rates of people in moderate and severe food insecurity have been flat and are expected even to increase slightly.  Overall, the world has made steady but slow progress on reducing the rate of food insecurity.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/Rates.pdf}
  \caption{Percentage of the population over the threshold for moderate and severe food insecurity, by World Bank income groups as of 2020}
  \label{fig:rates}
\end{figure}

\subsection{Food Insecurity Across Space}
We find substantial heterogeneity in the global distribution of severe food insecurity across countries and subnational regions (See Fig. \ref{fig:map}).  For the year 2020, mainland sub-Saharan Africa is the continent with the highest rates of severe food insecurity, with 20\% of people over the corresponding threshold in at least one subnational area in every country except Gabon, Djibouti and Equatorial Guinea.  Outside of Africa, serious pockets of severe food insecurity also occur in Venezuela, Syria, Papua New Guinea, Yemen, and Afghanistan.  In many large middle-income countries, severe food insecurity is also quite prevalent, with rates between 10\% and 15\% in northern Brazil, many central Asian and middle-eastern countries, as well as India and Indonesia.

The experience of at least moderate food insecurity is quite common in many parts of the world.  In 2020, much of Africa, south and southeast Asia, and parts of Latin America had over 50\% of the population living with moderate food insecurity.  Even in highly developed economies, such as the United States and parts of eastern and southern Europe, over 10\% of the population is above the threshold for moderate food insecurity.

\begin{landscape}
\begin{figure}[h]
  \centering
  \includegraphics{img/FullMap.png}
  \caption{Spatial distribution of moderate and severe food insecurity for the years 2010, 2020, and 2030.}
  \label{fig:map}
\end{figure}
\end{landscape}

\subsection{Important Predictors of Food Insecurity}
Overall, our model was able to predict food insecurity with high accuracy in the countries for which training data was available ($R^2 = 0.950$ and $R^2 = 0.944$ for moderate and severe food insecurity, respectively), and excluding entire countries from the training data to evaluate out-of-country model error still showed overall good results ($R^2 = 0.799$ and $R^2 = 0.772$ for moderate and severe food insecurity, respectively).  The relative importance of different variables as predictors of food insecurity can be assessed by examining the loss in forecast accuracy resulting from the exclusion of particular covariates.

For severe food insecurity, the most important variable for predicting food security levels is the Poverty Headcount Index, followed by the rate of stunting.  For predicting the rate of people with at least moderate food insecurity, GDP per capita is the most important, followed closely by the Poverty Headcount Index.  For the rate of severe food insecurity, the country's Gini coefficient was more important compared to moderate food insecurity.  Across both indicators, economic variables were the most relevant predictors.

@@

\section{Discussion}
This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts until the year 2030.  We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the coronavirus pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in developing countries.  Nevertheless, lower and lower-middle income countries, are expected to see overall increases in the number of food insecure people.

Our analysis of the relative importance of different predictors for future food insecurity dynamics shows that economic conditions are the largest driver of global variation in rates of food insecurity. For both moderate and severe food insecurity, the Poverty Headcount Index, the Gini coefficient, and GDP per capita appear as highly relevant predictors of the rate of food insecurity. The rate of stunting, an indicator of both sanitary conditions and chronic, long-term hunger is also an important predictor of food insecurity. However, other indicators of sanitary conditions, such as the rates of wasting and malaria mortality are less important as explanatory factors of changes in food insecurity. Geographic and climate factors, such as the percentage of urban population, topographic ruggedness and mean annual temperature or precipitation do not appear to be robust predictors of food insecurity at a global scale. Finally, while economic conditions are widely associated with food insecurity at a global scale, the fact that high income countries have seen little improvement in food insecurity outcomes, in spite of marginal annual increases in GDP per capita, suggests that there is an upper limit to the extent to which increasing economic development can improve food security outcomes.

Our results anticipating of increases in the number of people over the threshold for moderate food insecurity concur with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.  However, severe food insecurity may be a more apt indicator to compare with undernourishment, as the global totals are more similar \citep{sofi2020}.  In this case, our predictions diverge: the FAO estimates a 24\% increase, from 687 to 841 million undernourished people, from 2019 to 2030, whereas we predict a 3\% decrease, from 788 to 767 million severely food insecure people over the same period.

These discrepancies are in part due to our differing modeling approaches. While the SOFI report extrapolated recent trends in undernourishment and stunting, we modeled food insecurity based on interactions between projections in factors like GDP per capita, population, and the prevalence of stunting. Additionally, our diverging predictions for different hunger indicators are not necessarily mutually exclusive: it is theoretically possible for undernourishment to increase while severe food insecurity, as measured by the FIES, decreases.  The discrepancies between our projections and those of the FAO illustrate both the complexity of hunger as a social phenomenon, as well as the challenges in predicting future socioeconomic developments, where different modeling approaches may yield different forecasts.

When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and hunger are the same across all locations.  We test the accuracy of these cases by using ten-fold cross validation at the country level, and we find reasonable accuracy ($R^2 = 0.799$, MAE = 0.068 for moderate food insecurity and $R^2 = 0.772$, MAE = 0.038 for severe food insecurity).  This suggestions that our model explains between 77\% to 79\% of the variation in food security outcomes and is on average, within 3-7 percentage points in its estimates of rates of food insecurity, when applied to countries outside of the training set.

Additionally, when making predictions into the future, we assume that the long-term patterns of demographic change, urbanization, and development will maintain their trajectories for the next decade.  At a global scale, this is almost certainly the case: it is highly unlikely that rates of fertility or economic growth will shift suddenly and globally.  Nevertheless, at more local scales, sudden crises can lead to severe and unforeseeable increases in food insecurity, as crises in the previous decade in Yemen, Syria and Venezuela have shown.

While we find that rates of food insecurity are expected to decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, still expects the number of people experiencing moderate food insecurity to increase and the number of people experiencing severe food insecurity to fall only slightly between 2010 and 2030.  This still leaves billions of people eating less than they should and three-quarters of a billion people going entire days without eating a decade from now, a falling far short of the SDG 2 goal of ending hunger.  Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, sustainable diets, or obesity.  Thus, while expected trends in correlates of food security give cause for optimism, there is still significant work to be done at the political level in pursuit of SDG 2.

\section{Methods}
\subsection{Disaggregation}
The data on the FIES collected by Gallup records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.  Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.  We use a similar methodology and calculate post-stratification weights at a subnational level.  We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.  For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}, where available.  This methodology rests on the mild assumption that subnational differences in rates of food insecurity are driven by factors related to demographics, urbanization, and wealth rather than by other unobservable factors.

\subsection{Covariates}
We model subnational rates of severe and moderate-to-severe food insecurity as a function of several covariates related to human development, food security, health, infrastructure, income levels and distribution, and the environment (See Table \ref{tab:covars}).  We employ covariates for which projections are available up to year 2030 and, in most cases, also have subnational spatial resolution.  For projections based on the Shared Socioeconomic Pathways (SSP) framework \citep{oneill2014new}, we use the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on the Representative Concentration Pathway (RCP) 6.0 \citep{van2011representative}.

Many of the variables require the harmonization of subnational historical data with projections at the national level.  These include mean years of schooling, GDP per capita, as well as population. We use observed trajectories in the historical distribution of variables among subnational areas within a country to predict the future distribution of GDP, population, and schooling among subnational areas and disaggregate national-level future projections.

For health variables that have historically shown long-term trends and represent a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these and to obtain predictions up to variables for 2030. We calculate the rate of change between each pair of years in the dataset, and then use the mean rate of change over the period for which data are available, giving greater weight to more recent years, and applying that mean rate of change to estimate future levels.  This method is used to obtain forecasts of the rates of stunting and wasting up to the year 2030.

For the climatic variables, temperature and precipitation, we combine historical observations with an ensemble of four bias-corrected simulations of the future climate from the Inter-Sectoral Impact Model Intercomparison Project (ISIMIP) \citep{warszawski2014inter}.

Finally, we adjust several variables to account for the effects of the COVID-19 pandemic.  These include June 2020 GDP predictions from the World Bank \citep{prospects2020}, August 2020 estimates of impacts on anthropometry from \textit{The Lancet} \citep{headey2020impacts}, as well as the Poverty Headcount Index \cite{Cuaresma2018}.  For a detailed overview of the steps involved in processing and preparing each covariate, see the Appendix.

\begin{table}[H]
  \centering
	\begin{tabular}{lll}
		\toprule
    Name & Source & Scale \\
		\midrule
    Urban Percentage & \citep{Jones2016} & Subnational \\
    Stunting & \citep{Local2020} & Subnational \\
    Wasting & \citep{Local2020} & Subnational \\
    Mean Years of Schooling & \citep{Smits2019, KC2017} & Subnational\\
    GDP Per Capita & \citep{Smits2019, Dellink2017} & Subnational \\
    Gini Coefficient & \citep{Rao2019a} & National\\
    Poverty Headcount Index & \citep{Cuaresma2018} & National \\
    Water Scarcity & \citep{greve2018global} & Subnational \\
    Mean Annual Precipitation &  \cite{abatzoglou2018terraclimate, warszawski2014inter} & Subnational \\
    Topographic Ruggedness &  \cite{USGS1996, Riley1999} & Subnational \\
    Mean Temperature &  \cite{abatzoglou2018terraclimate, warszawski2014inter} & Subnational \\
    Malaria (\textit{P. falciparum}) Mortality Rate &  \cite{Weiss2019} & Subnational \\
		\bottomrule
	\end{tabular}
	\caption{Covariates Included in the Analysis}
	\label{tab:covars}
\end{table}

\subsection{Modeling}
We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.  We use random forest regressions, which perform well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.  A random forest regression involves generating a large number of decision trees, each using different sub-samples of the data, and then aggregating the predictions of the decision trees. Given the bounded nature of our variables of interest, the random forest regression are applied to the logistically transformed rates.
